---
title: "Analysing India's demonetisation and its potential impact on black money."
output:  html_document
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=4,scipen = 9)
library(xlsx)
library(dplyr)
library(tidyr)
library(xtable)
lakh<-100000;
crore<-100*lakh;
```
# Overview.
The source code and all calculations used in this analysis are available at https://github.com/DataDrivenPolicy/India.

There is a difference between unaccounted or untracked transaction and black money. 

* A farmer selling his crop and receiving returns in cash is not accumulating black money.
* Your maid and innumerable other members of the unorganized sector who receive payment in cash and whose annual income is less than the taxable limit are not generating black money.
* Even members of the business community earning crores through cash only transaction who keep proper accounts and pay the required taxes are not generating black money.

The revenue earned can be classified based on legitimacy of earning source and reporting of the earned revenue. 

e.g.

* A CM legitimizes kickback by routing it as donation to his families charitable trust.
* Another CM legitimizes kickback by manipulation of stocks of a family owned company and selling stocks at inflated rates.
* An MP receives 'legitimate' kickback by acquiring high value property and giving it on rent with a guarantee from the renter to pay off the bank loan via rent within 5 years.

All these transaction will be reported to the tax department and proper tax will be paid.

On the other hand we have legitimate transactions turned illegitimate due to non payment of taxes.
e.g.

* Someone selling off an acre of land to finance emergency healthcare, education or marriage expenses.
* A retired couple living off the rental income of their property not classifying it as leased property which will attract a higher property tax slab.

Typically when we speak of corruption we do not speak of the individual one off corruption where the earning is legitimate but taxes are not paid, we speak of the systemic corruption where the earnings are from a criminal and anti-social activity. We speak of the Sub Inspector, the Tehesildar, the politician the business magnet who have earned their wealth through illegal means and have stored crores of it in barns, wells, attics etc.

The demonetization scheme is supposed to render this physical cash invalid and provide practical hurdles in converting them to legitimate cash.

We will analyse the amount of physical cash which is used in regular transactions vs what is stored long term in an illegal fashion.

Note: While we would have preferred fine grained quantile data for the analysis, previous efforts to obtain this data via RTI have not been successful. So we will use the data "average" reported by various agencies and available in public domain 

```{r}
# Read the average data stored in an Excel Sheet.
df<-read.xlsx2("commonData.xlsx",1,colClasses = c("character","numeric","numeric","character","character"))

# Convert data to a common unit.
df1<-df%>% mutate(Value=Value*multiplier)

# Put the data in a named list.
cd<-as.list(df1$Value)
names(cd)<-df1$Item
```

# Understanding currencies in circulation.

As per data from the RBI there are  `r cd$amount/(lakh*crore)` lakh crore worth of currencies in circulation. This is NOT the total black money. Most of it is the cash we use to pay for petrol, milk, vegetables etc.

What "currency in circulation" means is that if we take all the currencies printed since Independence and subtract those currencies that were withdrawn or returned as soiled/mutilated and destroyed or are in the vaults of RBI or various banks, we get a value of Rs `r cd$amount/(lakh*crore)` lakh crore.

While this number looks big, it is not.

* If equally distributed across the `r cd$Population/crore` crore Indians, each Indian will receive `r cd$amount/(cd$Population)` Rupees. Far, Far less than the fabled 15 lakh rupees.

Most of the readers of this article must be carrying at least half of this amount in their pockets and may have withdrawn more than this amount at the start of the month to cover household expenses.

## Let us look at the basic survival needs.

NSSO declares the average monthly per capita expenses as `r cd$"per capita expenditure Annual"/12` 

If the reader is muttering that my monthly fuel/ola/bus bill is more than this, then pause to consider and digest the fact that you and your experiences do not represent the average Indian.

```{r}
emergencyTransactions<-cd$Population*cd$`per capita expenditure Annual`/12
```

This mythical average Indian would have withdrawn `r cd$"per capita expenditure Annual"/12` Rs from Bank or set aside from salary the required budget to meet his expenditure for the month of November. With demonetization announced at the start of the month he and others like him would rush to the banks to  maintain cash liquidity in effect the collective population would have transacted approximately Rs `r emergencyTransactions/(lakh*crore)` lakh crores after Nov 8th.

If this looks similar to what the government has reported as a sudden spurt in Jan Dhan accounts then it must be co-incidence. http://economictimes.indiatimes.com/news/politics-and-nation/government-looking-into-sudden-spurt-in-jan-dhan-account-deposits/articleshow/55390173.cms 

## Understanding the income side of the problem.

The average annual per capita income is Rs `r cd$"Per Capita Income Annual"`
The interval at which this income is recieved id different for each individual, on the rural side.

* A farmer growing sugarcane will be paid once in a year.
* Someone else growing kharif and rabi crops may get paid twice a year.
* Some one involved in Dairy farming may be paid daily or at whatever inverval decided by the milk cooperative.

Similarly on the industrial/commercial side

* A day wage earner or a shopkeeper will be paid daily
* A business man may be paid after product delivery which may take months.
* A salaried employee may be paid monthly


Now  only `r cd$"population with bankaccount %"` % of Indians have bank accounts of which `r cd$"dormant account percentage"` % of the accounts are dormant.
```{r}
activeAccount<-cd$"population with bankaccount %" * cd$"dormant account percentage"/100
```
So only `r activeAccount` % of the population have an active bank account.

While those with a bank account may deposit their income in their account `r 100 - activeAccount` % of the citizens do not have access to banking and may transact in cash only. Since we do not know the frequency of payout (daily,weekly, monthly, quarterly,half yearly, yearly) The total cash in hand can vary between
`r x<-(100-activeAccount)/100 * cd$Population * cd$"Per Capita Income Annual";x/(lakh*crore)` lakh crore if every one was paid on an annual basis and `r x/(lakh*crore*365)` lakh crore if the payment was on daily basis.

Some where between these two extremes, with on one hand the public holding `r x/cd$amount` times more cash than what is available in the system and on the other hand the public holding cash just enough for their bare necessity lies teh actual cash situation. We have people at both the two extremes, but the average lies somewhere in the middle, In the absence of any data on where the average lies, let us assume that 

* The Cash only community gets paid Rs `r incomeMonthly<-cd$"Per Capita Income Annual"/12; incomeMonthly` once in a month.
* Since they do not have access to banking they store the left over cash from the previous months as follows.
    *  Cash equivalent to one months income @home in the rice bin or pooja room. for emergency expenses.
    *  A percentage every month in unregulated investment schemes like the local chit funds/gold schemes/money lenders etc.
    *  Rest in assets like property, land, gold, motor vehicles, Gadgets etc.

So the cash in hand would be Rs `r 2* incomeMonthly` i.e out of Rs `r cd$amount/(lakh*crore)` lakh crore rupees circulating in the economy, `r 2 * incomeMonthly*cd$Population*(100-activeAccount)/(lakh*crore*100)` lakh crore rupees is held as cash of `r 2*incomeMonthly ` by `r (100-activeAccount)`% of the population without access to banking.

`r remaining<-(cd$amount- (2 * incomeMonthly*cd$Population*(100-activeAccount)/100))`
Out of the remaining Rs `r remaining/(lakh*crore)` lakh crore some is legitimately held as monthly expenses by those with access to banking sector i.e. Rs `r (cd$Population*activeAccount/100) * cd$"per capita expenditure Annual"/(lakh*crore*12)` lakh crore and as petty cash (minimum of `r cd$"Active Companies"*cd$"petty cash Min"/(lakh*crore)` lakh crore by business.
`r remaining<-remaining - (cd$Population*activeAccount/100) * cd$"per capita expenditure Annual"/12 - cd$"Active Companies"*cd$"petty cash Min" `

While there may be other reasons of holding legitimate cash (marriage, medical emergency, organizing events, house construction etc) It will be difficult to estimate this amount, also even if these cash holdings are large in value we do not expect it to be held by large enough sections of the population to create more than a percentage point difference in our calculations.

Our Estimate for unaccounted cash holding is Rs. `r remaining/(lakh*crore)` lakh crore.
i.e. only `r 100*remaining/cd$amount` % of the floating cash may be hoarded in cash format.

cd$"
---
The average annual per capita income for an average Indian is Rs `r cd$"Per Capita Income Annual"`
i.e. the combined income of all Indians is `r cd$Population * cd$"Per Capita Income Annual"/(lakh*crore)` lakh crore. Even if we assume that every bank account is a salary account and the rest of the population get their income as cash, then the cash payout is  `r cd$Population * cd$"Per Capita Income Annual" *(100-activeAccount)/(100*lakh*crore)` lakh crore rupees p.a. 

If we look at only the expenditure number, then on the first of every month every citizen get's ready to spend Rs `r cd$"per capita expenditure Annual"/12` resulting in a total of `r cd$Population* cd$"per capita expenditure Annual"/(lakh*crore*12)` lakh crore Rs. being kept ready to meet this months expenses. Since the demonetization scheme was launched in the initial part of the week, most of this cash is still unspent and has to be converted to the new currency ASAP to ensure household financial  liquidity.

```{r}
annualPaymentsNonBanking<- cd$Population * cd$"Per Capita Income Annual" *(100-activeAccount)/(100*lakh*crore)
annualExpenditureNonBanking<- cd$Population * cd$"per capita expenditure Annual" *(100-activeAccount)/(100*lakh*crore)
annualSavingsNonBanking=annualPaymentsNonBanking-annualExpenditureNonBanking
```
The cash payout as salary occurs in the first week of every month. So a total of `r annualPaymentsNonBanking/12` lakh crore rupees is earned by the average non-banking Indian every month. Which also indicates that this amount cannot be locked up as black money as it has to be in regular circulation.

If we look at the monthly expenditure and calculate the savings that will be carried over in cash form by this section we get `r (annualPaymentsNonBanking-annualExpenditureNonBanking)/12` lakh crores saved every month. If this sector decides to keep their saving in cash format, then the currency in circulation will be locked up in `r (cd$amount/(lakh*crore) - (annualPaymentsNonBanking/12))/(annualSavingsNonBanking/12)` months.

The remaining `r activeAccount`% would withdraw at least cash equal to their monthly needs from the bank.
NSSO survey indicates that the monthly expenditure per person is `r cd$"per capita expenditure Annual"/12` resulting in a sum of `r activeAccount*cd$Population*cd$"per capita expenditure Annual"/(100*12*lakh*crore)` lakh crore being withdrawn from the banks monthly.

The average annual expenditure for an Indian is Rs `r cd$"per capita expenditure Annual"`

```{r}
avgSavings<- cd$"Per Capita Income Annual"- cd$"per capita expenditure Annual"
```



The annual savings are `r avgSavings*cd$Population /cd$amount` times more than the cash available in the system

Given that it is not possible to store money in cash format for more than two months without creating a currency crisis. The excess savings every year needs to be converted in some other form.

The banks have declared an additional deposit of only Rs `r cd$"annual FD Increase"/(lakh*crore)` lakh crore in this financial year

```{r}
popNonBankingSavings<-((100-activeAccount)/100) * avgSavings*cd$Population/(lakh*crore)
```

48% of the population is engaged in agriculture where the revenue will be available either once, twice or thrice an year depending on the crop being grown and the availability of irrigation. So even in the best case this section has to hold 4 months worth of liquid cash in reserve.
Assuming that a person without access to banking will maintain cash liquidity for emergencies equal to at least a month expenditure.
```{r}
reserveCash<-(.48*4+.52*1)*cd$"per capita expenditure Annual"/12
```
resulting in a reserve cash of Rs `r reserveCash` per person or `r reserveCash *cd$Population/(lakh*crore)` nation wide.

note: the numbers used in these calculation and their source are as follows
```{r results="asis"}
print(xtable(df),type="html")
```

