---
title: "Banananomics"
output:
  html_document: 
    self_contained: no
  html_notebook: default
---
```{r setup}

rm(list=ls())
library(XML)
library(dplyr)
library(tidyr)
library(lubridate)
library(rgdal)
library(leaflet)
library(xtable)
library(ggplot2)
knitr::opts_knit$set(root.dir = normalizePath('../'))
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE)
options(digits=4,scipen = 9)
```
There are reports of various informal markets collapsing due to demonetisation. 
A market can collapse due to one of the following factors.

1. Fall in Supply.
2. Fall in Demand.
3. Logistic problems which prevent connection of supply to demand.

A collapse in market should result in either low volume of goods sold or if the volume is constant then it should result in lower prices for the goods.

An increase in price can indicate either low supply or a high demand. Similarly a decrease in price can indicate either a high supply or a low demand.

And finally an increase and/or decrease can also indicate lack of a marketplace for supply and demand to meet.
```{r cleanup , cache=TRUE}
getwd()
if(!file.exists("data/Banana_2016.xml")){
  download.file("https://data.gov.in/sites/default/files/Banana_2016.xml",destfile = "data/Banana_2016.xml")
}
x<-xmlParse("data/Banana_2016.xml")
y<-getNodeSet(x,'//Table')
z<-xmlToDataFrame(y)
```
```{r}
a<-z%>%
  mutate(Date=dmy(Arrival_Date))%>%
  filter(Date>dmy("01/10/2016"))%>%
  mutate(Week=as.numeric(Date-dmy("01/11/2016"))%/%7)%>%
  mutate(Week=paste0("Week",Week),Price=extract_numeric(Modal_x0020_Price))%>%
group_by(District,Variety,Week)%>%
 summarise(mp=mean(Price)) %>%
  spread(Week,mp)%>%
  mutate(percentage=Week0)%>%
  gather(weeks,Average,starts_with("Week"))
a<-a[complete.cases(a$percentage),]

    b<-a%>%
      mutate(percentage=100*Average/percentage)%>%
      ungroup()%>%
  group_by(District,weeks)%>%
  summarise(meanPercentage=mean(percentage))%>%
      spread(weeks,meanPercentage)
c<-b%>%gather(Weeks,Percentage,starts_with("Week"))%>%ungroup()%>%group_by(Weeks)
c<-c[complete.cases(c),]
d<-c%>%summarise(Mean=mean(Percentage),Median=median(Percentage),Min=min(Percentage),Max=max(Percentage))
ggplot(c,aes(Percentage,fill=Weeks))+geom_histogram(position="dodge")+facet_wrap(~Weeks)
```

```{r,results="asis"}

print(xtable(d,caption ="Weekly Percentage Price movement"),type="html")
```

```{r map}
shp<-readOGR("../maps/Districts/Census_2011")
shp@data<-left_join(shp@data,b,by=c("DISTRICT"="District"))
shp@data<-shp@data%>%mutate(popup=paste0("<b>District=",DISTRICT,"</b><br/>","Week 1=",Week1, "<br/>Week2=",Week2, "<br/>Week3=",Week3))
pal<-colorBin(rev(topo.colors(10)),shp$Week3,na.color="#ffffff")
leaflet(shp)%>%
addTiles(urlTemplate = "http://{s}.basemaps.cartocdn.com/light_all/{z}/{x}/{y}.png") %>%
addPolygons(stroke=F,fillColor =pal(shp$Week3), popup=shp$popup)  %>%
addLegend(position = 'topright', pal=pal,values=shp$Week3Percentage, opacity = 1.0,       title = 'Banana Wholesale rate Nov 2 to Nov 28')

```